---
title: "The Noisy Work of Uncertainty Visualisation"
author: 
  - name: Harriet Mason
    url: https://harrietmason.netlify.app/
    orcid: 0009-0007-4568-8215
    email:  harriet.m.mason@gmail.com
    affiliation:
    - name: Monash University
      department: Department of Econometrics and Business Statistics
      city: Melbourne
      country: Australia
  - name: Dianne Cook
    url: https://dicook.org
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu
    affiliation:
    - name: Monash University
      department: Department of Econometrics and Business Statistics
      city: Melbourne
      country: Australia
  - name: Sarah Goodwin
    url: https://www.linkedin.com/in/smgoodwin/
    orcid: 0000-0001-8894-8282
    email: Sarah.Goodwin@monash.edu
    affiliation:
    - name: Monash University
      department: Department of Human Centred Computing
      city: Melbourne
      country: Australia
  - name: Emi Tanaka
    url: https://emitanaka.org/
    orcid: 0000-0002-1455-259X
    email: Emi.Tanaka@anu.edu.au
    affiliation: 
    - name: The Australian National University
      department: Biological Data Science Institute
      city: Canberra
      country: Australia
  - name: Susan VanderPlas
    url: https://srvanderplas.github.io
    orcid: 0000-0002-3803-0972
    affiliation:
    - name: Utah State University
      department: Department of Mathematics & Statistics
      city:  Logan, UT
      country: United States
bibliography: paper.bib
abstract: Better representation of the uncertainty in a data visualisation is a focus of recent research activity. A problem with the current literature is that there is a lack of clarity about the definition of uncertainty and what it means to represent it in a plot. This confusion results in a significant amount of conflicting results in the literature, especially in experiments that assess the effectiveness of different uncertainty representations. In this review, we summarise the current literature, provide workable definitions, and illustrate these definitions with examples. In doing so, we ask what it really takes to achieve transparency in statistical graphics. It is hoped that it will be useful for guiding new graphics methodology and experimental research.
date: last-modified
toc: false
number-sections: true
latex-clean: true
format:  
  jasa-pdf:
    keep-tex: true  
    journal:
      blinded: false
    filters:
    - at: pre-quarto
      path: _extensions/andrewheiss/wordcount/citeproc.lua
    - at: pre-quarto
      path: _extensions/andrewheiss/wordcount/wordcount.lua

  jasa-html: 
    filters:
    - at: pre-quarto
      path: _extensions/andrewheiss/wordcount/citeproc.lua
    - at: pre-quarto
      path: _extensions/andrewheiss/wordcount/wordcount.lua
fig-valign: bottom
cap-location: bottom
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| message: false
#| warning: false


# load Libraries
library(tidyverse)
# devtools::install_github("lydialucchesi/Vizumap")
library(RColorBrewer)
library(scales)
library(sf)
# devtools::install_github("UrbanInstitute/urbnmapr")
library(urbnmapr)
library(flextable)
library(colorspace)
#library(rgeos)
library(ggthemes)
library(distributional)
library(patchwork)
library(ggdist)
library(ggdibbler)
```

```{r}
#| eval: false
#| echo: false
# Get map data: do this once and save
my_map_data <- get_urbn_map("counties", sf = TRUE) |>
  filter(state_name=="Iowa")
save(my_map_data, file="data/iowa_map.rda")
```

```{r}
#| eval: false
#| echo: false
# Get centroid because rgeos is depreciated
load("data/iowa_map.rda")
centroids <- as_tibble(gCentroid(as(my_map_data$geometry, "Spatial"), byid = TRUE))
my_map_data$cent_long <- centroids$x
my_map_data$cent_lat <- centroids$y
save(my_map_data, file="data/my_map_data.rda")
```

```{r mapdata}
#| echo: false
#| message: false
#| warning: false

# get data
load("data/my_map_data.rda")

# seed for sampling
set.seed(1997)

# data dimension for sampling
n <- dim(my_map_data)[1]

# Make palettes
longpal <- rev(sequential_hcl(13, palette = "YlOrRd"))
basecols <- longpal[3:10]
breaks <- 21:29 
breakslong <- 18:32 
names(basecols) <- seq(8)
names(longpal) <- -1:11
var_seq <- seq(0,12, by=3)

my_map_data <- my_map_data |>
  mutate(temp = 29 - 2*abs(scale(cent_lat) - sin(2*(scale(cent_long)))[,1]), # trend
         highvar = runif(n, min=8, max=var_seq[5]), # high variance
         lowvar = runif(n, min=2, max=4), # low variance
         count_id = row_number()) |>
  pivot_longer(cols=highvar:lowvar, names_to = "variance_class", values_to = "variance") |>
  # add bivariate classes to data
  mutate(bitemp = cut(temp, breaks=breaks, labels=seq(8)),
         bivar = cut(variance, breaks=var_seq, labels=seq(4)),
         biclass = paste(bitemp, bivar, sep="-")) |>
  mutate(highlight = ifelse(count_id <= 5, TRUE, FALSE)) |>
  mutate(temp = as.numeric(temp))

```

```{r example-regressions}
#| echo: false
#| message: false
#| warning: false
# Linear regression - Case 1
# fit linear regression
fit1 <- lm(mpg ~ wt, data = mtcars)
model_summary1 <- summary(fit1)

# pull out estimates
int_est_1 <- coef(fit1)[1]
int_se_1 <- model_summary1$coefficients["(Intercept)", "Std. Error"]
slope_est_1 <- coef(fit1)[2]
slope_se_1 <- model_summary1$coefficients["wt", "Std. Error"]

# Linear regression - Case 2
# sample data
set.seed(14) # selected seed to make spurious relationship
small_cars <- mtcars |>
  sample_n(3) 
# fit regression
fit2 <- lm(mpg ~ wt, data = small_cars)
model_summary2 <- summary(fit2)

# pull out estimates
int_est_2 <- coef(fit2)[1]
int_se_2 <- model_summary2$coefficients["(Intercept)", "Std. Error"]
slope_est_2 <- coef(fit2)[2]
slope_se_2 <- model_summary2$coefficients["wt", "Std. Error"]

# Estimate Plots
reg1a <- mtcars |>
  ggplot(aes(wt, mpg)) + 
  geom_point(size = 2, alpha = 0.5) + 
  theme_few()  + 
  geom_abline(intercept = int_est_1,
              slope = slope_est_1,
              colour = "dodgerblue3",
              linewidth = 1.5) +
  theme(aspect.ratio=1) +
  ggtitle("R1 - low") + 
  ylim(9,35) +
  xlim(1.5,5.5)

reg1b <- small_cars |>
  ggplot(aes(wt, mpg)) + 
  geom_point(size = 2, alpha = 0.5) + 
  theme_few()  + 
  geom_abline(intercept = int_est_2,
              slope = slope_est_2,
              colour = "dodgerblue3",
              linewidth = 1.5) +
  ggtitle("R2 - high") + 
  theme(aspect.ratio=1) +
  ylim(9,35) +
  xlim(1.5,5.5)

# Stat plot
reg2a <- mtcars |>
  ggplot(aes(wt, mpg)) + 
  geom_point(size = 2, alpha = 0.5) + 
  theme_few()  + 
  geom_abline(intercept = int_se_1,
              slope = slope_se_1,
              colour = "dodgerblue3", 
              linewidth = 1.5) +
  ggtitle("R1 - low") + 
  theme(aspect.ratio=1) +
  ylim(0,35) +
  xlim(0,5.5)

reg2b <- small_cars |>
  ggplot(aes(wt, mpg)) + 
  geom_point(size = 2, alpha = 0.5) + 
  theme_few()  + 
  geom_abline(intercept = int_se_2,
              slope = slope_se_2,
              colour = "dodgerblue3",
              linewidth = 1.5) +
  ggtitle("R2 - high") + 
  theme(aspect.ratio=1) +
  ylim(0,35) +
  xlim(0,5.5)


# Width plot
reg3a <- mtcars |>
  ggplot(aes(wt, mpg)) + 
  geom_point(size = 2, alpha = 0.5) + 
  theme_few()  + 
  geom_abline(intercept = int_est_1,
              slope = slope_est_1,
              alpha = 2/int_se_1,
              linewidth = 2*slope_se_1,
              colour = "dodgerblue3") +
  ggtitle("R1 - low") + 
  theme(aspect.ratio=1) +
  ylim(9,35) +
  xlim(1.5,5.5)

reg3b <- small_cars |>
  ggplot(aes(wt, mpg)) + 
  geom_point(size = 2, alpha = 0.5) + 
  theme_few()  + 
  geom_abline(intercept = int_est_2,
              slope = slope_est_2,
              alpha = 2/int_se_2,
              linewidth = 2*slope_se_2,
              colour = "dodgerblue3") +
  ggtitle("R2 - high") + 
  theme(aspect.ratio=1) +
  ylim(9,35) +
  xlim(1.5,5.5)


# Sample plot
reg4a <-  mtcars |>
  ggplot(aes(wt, mpg)) + 
  geom_point(size = 2, alpha = 0.5) + 
  theme_few()  + 
  geom_abline_sample(intercept = dist_normal(int_est_1, int_se_1),
                     slope = dist_normal(slope_est_1,slope_se_1),
                     alpha=0.2, times=100,
                     colour = "dodgerblue3")+
  ggtitle("R1 - low") + 
  theme(aspect.ratio=1) +
  ylim(9,35) +
  xlim(1.5,5.5)

reg4b <- small_cars |>
  ggplot(aes(wt, mpg)) + 
  geom_point(size = 2, alpha = 0.5) + 
  theme_few() + 
  geom_abline_sample(intercept = dist_normal(int_est_2, int_se_2),
                     slope = dist_normal(slope_est_2, slope_se_2),
                     alpha=0.5, times=100,
                     colour = "dodgerblue3")+
  ggtitle("R2 - high") + 
  theme(aspect.ratio=1) +
  ylim(9,35) +
  xlim(1.5,5.5)


```

```{r example-univariate}
#| echo: false
#| message: false
#| warning: false
# Point with y axis
group_df <- tibble(x = c(1,2,3,4,5),
       se_low = c(1,1,1,1,1),
       se_high = c(1,2,3,4,5),
       group = c("A", "B", "C", "D", "E")) |>
  mutate(dist_low = x + dist_beta(1/se_low, 1/se_low),
         dist_high = x*dist_beta(1/se_high, 1/se_high))


# Estimate
uni1a <- group_df |>
  ggplot() +
  geom_point(aes(x=mean(dist_low), y=group), size = 3, colour = "forestgreen") +
  theme_few() + 
  ggtitle("G1 - low") + 
  labs(y = "treatment group",
       x = "treatment effect") + 
  theme(aspect.ratio=1)

uni1b <- group_df |>
  ggplot() +
  geom_point(aes(x=mean(dist_high), y=group), size = 3, colour = "forestgreen") +
  theme_few() + 
  ggtitle("G2 - high") + 
  labs(y = "treatment group",
       x = "treatment effect") + 
  theme(aspect.ratio=1)

# Statistic
uni2a <- group_df |>
  ggplot() +
  geom_point(aes(x=round(variance(dist_low), 4), y=group), size = 3, colour = "forestgreen") +
  theme_few() + 
  ggtitle("G1 - low") + 
  labs(y = "treatment group",
       x = "variance of treatment effect") + 
  theme(aspect.ratio=1)

uni2b <- group_df |>
  ggplot() +
  geom_point(aes(x=variance(dist_high), y=group), size = 3, colour = "forestgreen") +
  theme_few() + 
  ggtitle("G2 - high") + 
  labs(y = "treatment group",
       x = "variance of treatment effect") + 
  theme(aspect.ratio=1)

# Error bars
uni3a <- group_df |>
  ggplot(aes(xdist = dist_low, y=group)) +
  theme_few() +
  stat_pointinterval(.width = c(.95), colour = "forestgreen") + 
  ggtitle("G1 - low") + 
  labs(y = "treatment group",
       x = "treatment effect") + 
  theme(aspect.ratio=1)

uni3b <- group_df |>
  ggplot(aes(xdist = dist_high, y=group)) +
  theme_few() +
  stat_pointinterval(.width = c(.95), colour = "forestgreen") + 
  ggtitle("G2 - high") + 
  labs(y = "treatment group",
       x = "treatment effect") + 
  theme(aspect.ratio=1)

# Distribution
uni4a <- group_df |>
  ggplot(aes(xdist = dist_low, y=group)) +
  theme_few() +
  stat_dots(fill = "forestgreen", colour=NA) + 
  ggtitle("G1 - low") + 
  labs(y = "treatment group",
       x = "treatment effect") + 
  theme(aspect.ratio=1)

uni4b <- group_df |>
  ggplot(aes(xdist = dist_high, y=group)) +
  theme_few() +
  stat_dots(fill = "forestgreen", colour=NA) + 
  ggtitle("G2 - high") + 
  labs(y = "treatment group",
       x = "treatment effect") + 
  theme(aspect.ratio=1)

```

```{r example-maps}
#| echo: false
#| message: false
#| warning: false

###################### CHOROPLETH MAP ###################### 
choro1 <- my_map_data |>
  filter(variance_class=="lowvar") |>
  ggplot() +
  geom_sf(aes(fill = temp, 
              geometry = geometry), colour=NA) + 
  scale_fill_gradientn(colours = basecols) +
  #theme_void() +
  labs(fill = "temperature") +
  theme_map() +
  theme(legend.position = "inside", 
        legend.position.inside = c(0.1, -0.2),
        legend.direction = "horizontal",
        legend.background = element_rect(fill=NA),
        plot.title = element_text(size = 14)) + 
  ggtitle("S1 - low") 

choro2 <- choro1 %+% filter(my_map_data, variance_class=="highvar") +
  ggtitle("S2 - high") 

######################  Exceed Prob Map ######################  

exeed_data <- my_map_data |> 
  as_tibble() |>
  mutate(xprob = 1- pnorm(27, mean=temp, sd=variance)) 
             
exceed1 <- exeed_data |>
  filter(variance_class=="lowvar") |>
  ggplot() +
  geom_sf(aes(fill = xprob, 
              geometry = geometry), colour=NA) + 
  scale_fill_gradientn(colours = basecols, limits=c(0,1)) +
  #theme_void() +
  labs(fill = "P(X>27)") +
  theme_map() +
  theme(legend.position = "inside", 
        legend.position.inside = c(0.1, -0.2),
        legend.direction = "horizontal",
        legend.background = element_rect(fill=NA),
        plot.title = element_text(size = 14)) +   
  ggtitle("S1 - low") 

exceed2 <- exceed1 %+% filter(exeed_data, variance_class=="highvar") + 
  ggtitle("S2 - high") 

####################### Bivariate maps ###################### 

# Make bivariate palette
# Function to devalue by a certain amount
colsupress <- function(basecols, hue=1, sat=1, val=1) {
    X <- diag(c(hue, sat, val)) %*% rgb2hsv(col2rgb(basecols))
    hsv(pmin(X[1,], 1), pmin(X[2,], 1), pmin(X[3,], 1))
}

# recursively decrease value
v_val = 0.5
bivariatepal <- c(basecols,
                  colsupress(basecols, sat=v_val),
                  colsupress(colsupress(basecols, sat=v_val), sat=v_val),
                  colsupress(colsupress(colsupress(basecols, sat=v_val), sat=v_val), sat=v_val))
# establish levels of palette
names(bivariatepal) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")

# mapping
bivar1 <- my_map_data |>
  filter(variance_class=="lowvar") |>
  ggplot() +
  geom_sf(aes(fill = biclass, geometry = geometry), colour=NA) + 
  scale_fill_manual(values = bivariatepal) +
  theme_void() + 
  theme(legend.position = "none") +
  ggtitle("S1 - low") 
  
bivar2 <- bivar1 %+% filter(my_map_data, variance_class=="highvar") +
  ggtitle("S2 - high") 


bivar_pal<- expand_grid(bitemp = seq(1,8), bivar = seq(1,4)) |>
  mutate(biclass = paste(bitemp, bivar, sep="-"))

bivar3 <- bivar_pal |>
  ggplot(aes(x=bitemp, y=bivar, fill=biclass)) +
  geom_tile() +
  scale_fill_manual(values = bivariatepal) +
  theme_classic() +
  theme(aspect.ratio=0.2) + 
  scale_x_continuous(breaks = c(1,2,3,4,5,6,7,8),
                     expand=c(0,0)) +
  scale_y_reverse(breaks = c(0,1,2,3,4),
                  labels = c(0,3,6,9,12)) + 
  labs(x = "temperature",
       y = "SE") +
  theme(legend.position = "none",
        axis.line = element_blank(),
        axis.ticks = element_blank()) 

bivar1 <- bivar1 / bivar3
bivar2 <- bivar2 / bivar3

####################### VSUP maps ###################### 
# Function to combine colours for VSUP
colourblend <- function(basecols, p_length, nblend) {
    X <- rgb2hsv(col2rgb(unique(basecols)))
    v1 <- X[,seq(1,dim(X)[2], 2)]
    v2 <- X[,seq(2,dim(X)[2], 2)]
    if("matrix" %in% class(v1)){
      # hue issue wrap around pt 1
      v3 <- (v1+v2)
      v3["h",] <- ifelse(abs(v1["h",]-v2["h",])>0.5, v3["h",]+1, v3["h",])
      v3 <- v3/2
      # hue issue wrap around pt 2
      v3["h",] <- ifelse(v3["h",]>=1 , v3["h",]-1 ,v3["h",])
      hsv(rep(v3[1,], each=nblend), rep(v3[2,], each=nblend), rep(v3[3,], each=nblend))
      } else {
        v3 <- (v1+v2)
        v3["h"] <- ifelse(abs(v1["h"]-v2["h"])>0.5, v3["h"]+1, v3["h"])
        v3 <- v3/2
        v3["h"] <- ifelse(v3["h"]>=1 , v3["h"]-1 ,v3["h"])
        rep(hsv(h=v3[1], s=v3[2], v=v3[3]), p_length)
        }
}

VSUPfunc <- function(basecols, p_length, nblend){
  colourblend(colsupress(basecols, sat=0.5), p_length, nblend)
}

# VSUP
p = length(basecols)
VSUP <- c(basecols,
          VSUPfunc(basecols, p, 2),
          VSUPfunc(VSUPfunc(basecols, p, 2), p, 4),
          VSUPfunc(VSUPfunc(VSUPfunc(basecols, p, 2), p, 4), p, 8))

names(VSUP) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")

# VSUP maps
vsup1 <- my_map_data |>
  filter(variance_class=="lowvar") |>
  ggplot() +
  geom_sf(aes(fill = biclass, geometry = geometry), colour=NA) + 
  scale_fill_manual(values = VSUP) +
  theme_void() + 
  theme(legend.position = "none") + 
  ggtitle("a") 

fake_data <- my_map_data |>
  filter(variance_class=="lowvar") |>
  mutate(bivar = cut(variance, breaks=c(0,1,2,3,4), labels=seq(4)), 
         biclass = paste(bitemp, bivar, sep="-"))

vsup2 <- vsup1 %+% fake_data + 
  ggtitle("b") 


fake_data <- my_map_data |>
  filter(variance_class=="lowvar") |>
  mutate(bivar = cut(variance, breaks=c(0,1,2,3,4), labels=seq(4)),
         biclass = paste(bitemp, bivar, sep="-"))


vsup3 <- bivar3 + scale_fill_manual(values = VSUP)

vsup4 <- vsup3 + 
  scale_y_reverse(breaks = c(0,1,2,3,4),
                  labels = c(0,1,2,3,4))

vsup1 <- vsup1 / vsup3
vsup2 <- vsup2 / vsup4
####################### Pixel maps ######################
# low variance
pixel1 <- my_map_data |>
  filter(variance_class=="lowvar") |>
  ggplot() +
  geom_sf_sample(aes(fill = dist_normal(temp, variance), 
              geometry = geometry), colour=NA, times=30) + 
  #theme_void() + 
  scale_fill_gradientn(colours = longpal) +
  labs(fill = "temperature") +
  theme_map() +
  theme(legend.position = "inside", 
        legend.position.inside = c(0.1, -0.2),
        legend.direction = "horizontal",
        legend.background = element_rect(fill=NA),
        plot.title = element_text(size = 14)) +   
  ggtitle("S1 - low") 

# high variance
pixel2 <- my_map_data |>
  filter(variance_class=="highvar") |>
  ggplot() +
  geom_sf_sample(aes(fill = dist_normal(temp, variance), 
              geometry = geometry), colour=NA, times=30) + 
  scale_fill_gradientn(colours = longpal) +
  #theme_void() +
  labs(fill = "temperature") +
  theme_map() +
  theme(legend.position = "inside", 
        legend.position.inside = c(0.1, -0.2),
        legend.direction = "horizontal",
        legend.background = element_rect(fill=NA),
        plot.title = element_text(size = 14)) +   
  ggtitle("S2 - high") 
```


## Introduction
What do we mean when we talk about "uncertainty visualisation"?
The phrase can feel contradictory to anyone who is familiar with the term.
Among statisticians, "uncertainty" is often discussed as an omnipresent spectre touching every stage of our analysis without ever being fully seen. 
Authors will often mention that the phrase is vague [@Spiegelhalter2017; @Griethe2006], or avoid defining it by describing a list of things uncertainty *could* be [@Kinkeldey2014; @Hullman2016], but rarely do authors attempt to discuss what uncertainty actually *is*. 
<!--Almost as if in direct contrast with this, -->
By contrast, visual statistics (information visualisations, data plots) are one of the most powerful tools in the statistician's toolbox, allowing for quick and memorable communication that identifies quirks in our data that we didn’t even know to look for.
We see this in datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurus; @datasaurpkg], where visual statistics are able to highlight elements of the data that are invisible to the typical summary statistics.
We also see this in recall experiments, where simply sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].
Taken together, uncertainty visualisation implies a need to pull back the curtain and explore the unknowns of our analysis.

As nice as this sentiment is, it turns out to be easier said than done.
Reviews on uncertainty visualisation rarely offer tried and tested rules for effective uncertainty visualisation, instead commenting on the difficulties faced when trying to summarise the field.
@Kinkeldey2014 found most experimental methods to be ad hoc, with no commonly agreed upon methodology, formalisations, or a greater goal of describing general principles. 
@Hullman2016 noticed there is a serious noise issue in the field, with noise from participants misunderstanding visualisations, misinterpreting questions, and incorrectly applying heuristics overwhelming any information we can glean from studies.
@MacEachren2005 identified so much contradicting evidence that they spent an entire page discussing the conflicting evidence for the question "Should I map uncertainty to colour hue?"
@Spiegelhalter2017 concluded that different plots are good for different things, arguing against a universal best plot for all people and circumstances.
@uncertchap2022 summarised several cognitive effects that repeatedly arise in uncertainty visualisation experiments; these effects were each discussed in isolation as a list of considerations rather than an overarching theory for effective uncertainty visualisation.

<!--The problem faced by the literature is easily summarised with a famous quote by Henri Poincaré.-->

> "Science is built up of facts, as a house is built of stones; but an accumulation of facts is no more a science than a heap of stones is a house." - Henri Poincaré (1905)

While these reviews are thorough in scope, none discuss how the existing literature contributes to the broader goal of uncertainty visualisation -- that is, despite the wealth of reviews, the field of uncertainty visualisation remains a heap of stones. 
There is a mountain of work that identifies common heuristics found in uncertainty visualisations, evaluates competing plot designs, or starts a theoretical discussion on a niche aspect of the field. 
While important, each of these papers offers up its own bespoke motivation and methodology, with little reference to the uncertainty visualisation papers outside their fiefdom. 
<!--This becomes even more difficult to manage when these studies conflict. -->
The field is in desperate need of a unifying theory that can tie the conflicting and siloed research together. 
This review attempts to address this issue by offering a novel perspective on the uncertainty visualisation problem.
That is, we will use the wealth of established stone to construct a foundation to build a house.

## The purpose of uncertainty visualisation
Mentions of "uncertainty visualisation" start springing up around 1990, across several different fields [@Ibrekk1987; @MacEachren1992], each with its own motivation for the work.
In computer science, the area appears to be motivated by issues in the public's perception of random variables, with the hope that visualisations would give laypeople the ability to extract important information from graphical representations [@Ibrekk1987].
With similar concerns about the public's understanding of randomness, the fields of psychology, statistics, and economics used "uncertainty visualisations" as a communication tool to mitigate the psychological bias associated with the communication of risk, a topic of concern since the early 1980s [@Spiegelhalter2017].
In cartography, it was motivated by inherent uncertainty of geoscience data, the practical use of visualisation as an exploratory tool, and the constrained visual channels from map representations [@MacEachren1992].
These disparate motivations have blended together, and today, uncertainty visualisation is usually motivated by the vague goal of "decision-making".
This term has been used to mean the mitigation of psychological bias to ensure economically rational decisions [@Padilla2022; @Padilla2021; @Kale2021], the facilitation of trust or confidence [@Zhao2023; @Yang2023],  the ability to extract values related to a distribution [@Sarma2023], the prevention of false discovery in plots [@Sarma2024; @Koonchanok2023], or the extraction of some other metric that is vaguely related to "uncertainty" [@Chakraborty2024; @Ndlovu2023].
This gradual scope creep of the field, motivated by the hazy definition of "decision-making", is the most likely culprit for the jumbled literature that makes up the field today. 

Given that there is so much subliminal disagreement in uncertainty visualisation, how did these disparate motivations come to be seen as interchangeable? 
All discussions on uncertainty visualisation seem to have a common thread that connects them: the belief that the ultimate goal of uncertainty visualisation is not trust, rationality, or value extraction, but *transparency*.
We see it said directly in reviews of the field [@uncertchap2022], or when authors claim that failing to include uncertainty is akin to fraud or lying [@Hullman2020a; @Manski2020].
We see it when authors assert that uncertainty communicates the legitimacy (or illegitimacy) of the conclusion drawn from visual inference [@Correll2014; @Kale2018; @Griethe2006]. 
We see it when authors say uncertainty visualisations should communicate a degree of confidence [@Boukhelifa2012] or validity [@Hullman2020a; @Griethe2006] in our conclusions.
We see it when authors suggest uncertainty visualisation should "guide, qualify, or soften our judgements of uncertain data" (e.g. @Leland2005 in his seminal work on the grammar of graphics).
These authors are not wrong about the need for transparency in science communication: a six-month survey of anti-mask groups on Facebook during the COVID-19 pandemic showed that anti-maskers made persuasive arguments by exploiting inherent uncertainty ignored by pro-maskers [@Lee2021].

Uncertainty visualisation is motivated by the need for a sort of “visual hypothesis test”, a sentiment expressed by some authors directly [@Correll2018; @MacEachren1992]. 
A successful uncertainty visualisation would act as a “statistical hedge” for any inference we make using the graphic. 
Since the purpose of a visualisation is to give a quick gist of the information [@Spiegelhalter2017], this hedging should be communicated visually without the need for complicated mental calculations.
Therefore, an effective uncertainty visualisation should not just "show" uncertainty; untrustworthy conclusions should *not be visible*.
If we refer to the conclusion we draw from a graphic as its signal, where uncertainty should make this signal harder to read as the "noise" increases, we can summarise the above information into three key requirements. 
A good uncertainty visualisation should:

  1) Reinforce justified signals to encourage confidence in results.
  2) Hide spurious signals that are overwhelmed by noise.
  3) Perform tasks 1) and 2) in a way that is proportional to the level of confidence in those conclusions.

Usually, visualisations that are unconcerned with uncertainty have no issue showing justified signals, but struggle with the display of unjustified signals. 
Therefore, we **suggest calling this approach to uncertainty visualisation "signal-suppression"** since it primarily differentiates itself from the normal "noiseless" visualisation approach through criterion (2). 
This is the main criterion we will use to assess the current literature on uncertainty visualisation.

## Current Approaches
### Ignoring uncertainty
The most common way to visualise uncertainty is to simply not. 
A study conducted by @Hullman2020a found that only a quarter of authors surveyed included uncertainty in 50% or more of their visualisations, in part because authors are not sure how to calculate uncertainty [@Hullman2020a].
This is not entirely unreasonable, given that even visualisation authors themselves seem to be in conflict about what exactly uncertainty is. 
We will start with visualisations that ignore uncertainty with the hope that by looking at where uncertainty isn't, we can better understand where it is. 

#### What is uncertainty?
It is surprisingly hard to describe what uncertainty is.
Most authors avoid the problem and describe the many characteristics of uncertainty.
Often, uncertainty is split by factors such as whether it is due to true randomness or a lack of knowledge [@Begg2014; @Spiegelhalter2017; @Gustafson2019; @uncertchap2022; @Hullman2016; @utypo]; quantifiable or unquantifiable [@Spiegelhalter2017; @utypo; @uncertchap2022]; scientific or human [@Benjamin2018; @Gustafson2019]; systematic or random [@Sanyal2009]; statistical or bounded [@Gschwandtnei2016; @Olston2002]; accuracy or precision [@Griethe2006; @Benjamin2018; @Hullman2016]; etc.
There are enough qualitative descriptors of uncertainty to fill a
paper, but none of this is particularly helpful in understanding how to integrate it into a
visualisation. 

Rather than trying to define uncertainty by looking at the myriad ways in which it *does* appear in an analysis, we may find it easier to look at where it *does not*.
Descriptive statistics describe our sample as it is and summarise large data into a useable format, but they are not seen as the primary goal of modern statistics.
In 19th-century England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson). 
Practitioners of the approach believed statistics ended with descriptive statistics, as science must be based on actual experience and observations [@Otsuka2023]. 
In order to make statements about population statistics, future values, or new observations, we need to perform inference, which requires the assumption of the "uniformity of nature", that is, we need to assume that unobserved phenomena should be similar to observed phenomena [@Otsuka2023]. 
Positivists believed referencing the unobservable was bad science, embracing descriptive statistics due to the inherent certainty associated with them.
Since uncertainty is nonexistent in descriptive statistics, it is clear that uncertainty is a by-product of inference: uncertainty is the noise that is both inseparable from our inference and meaningless without it. 

If we consider uncertainty as a byproduct of statistical inference, it would make sense to simply quantify it as a distribution.
Suggesting distributions as a representation of uncertainty is not new.
@Kay2023 originally suggested thinking about uncertainty visualisations as visualisations with distribution inputs, to replace the commonly used mean and standard deviation, removing the assumption of a Gaussian distribution.
Taking this point a step further, we suggest that a distribution is sufficient to characterise *any* uncertainty, no matter how intangible.
While some researchers believe abstract notions of uncertainty, such as credibility [@Thomson2005], forecaster confidence [@Padilla2021], or uncertainty about uncertainty [@Hadjimichael2024], are too complex to be quantified, 
abstract concepts such as human belief or credibility are regularly quantified by Bayesians, and hierarchical approaches are often used to model uncertainty about uncertainty.

#### Example: ignoring uncertainty

If visualising uncertainty is fundamentally visualising a set of random variables, what does "ignoring" uncertainty look like?
@fig-ignore shows plots of data from three different scenarios, each with either high or low uncertainty. The plots show the expected value of the input distribution.
In plots R1 and R2, the expected value is the line from a simple linear regression on the `mtcars` data (available in `ggplot2` [@ggplot2], and originally from @mtcars), with differing sample size underlying the uncertainty. 
Plots S1 and S2 show a choropleth map of Iowa, where counties are coloured according to a simulated temperature measurement, with measurement error being the uncertainty associated with the measuring instrument. 
Plots G1 and G2 show samples simulated from five populations (A-E). In G1 distributions have the same low variance, and in G2 they have the same high variance.
In the linear regression, we see a downward trend, and because this is overlaid on the observed data we can get a sense of the uncertainty, which is not entirely ignored in this plot because the data is plotted. However, the uncertainty of interest relates to the regression line. In the map you should see a sine wave  spatial trend, and in the univariate distributions, an incremental increase in treatment.
If we were to ask a reader "Can you see a difference between the plots in the top row versus the bottom row?
Is the strength of the trend communicated through the visualisation?"
The answer to both of these questions will be no, for S1, S2, G1, G2, as the high and low variance cases are identical. For R1 and R2, the answer would likely only relate to the points in the plots, not the regression line.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ignore
#| fig-height: 6
#| fig-width: 10
#| out-width: 100%
#| fig-pos: t
#| fig-cap: "Three example types of data and associated plots that will be used to illustrate various choices of uncertainty representation throughout the paper: scatterplot and regression line (R1, R2), spatial choropleth map (S1, S2), grouped dotplot (G1, G2). Two levels of uncertainty (low, high) is used with each example. Only with plots R1 and R2, where the regression overlays the scatterplot, can we see a difference between low and high uncertainty. Ignoring uncertainty can lead to misrepresentation of data."

reg1a + choro1 + uni1a + reg1b + choro2 + uni1b + plot_layout(ncol=3, widths=c(30, 40, 30))

```

These examples will serve as touchstones for a discussion of uncertainty visualisation, focusing on the approaches suggested in the literature. 

### Uncertainty as a statistic
Uncertainty is also often treated as another statistic, for example, the  exceedance probability map of spatial data (see @Kuhnert2018 with software available in @Lucchesi2021), where the probability of exceeding a specified value is displayed on the map, placing the focus on extreme events. 
An additional example is the icon array used for risk communication (e.g. @Spiegelhalter2017), where probability is translated to a set of icons, like people images, coloured by outcome. The idea is that people will understand the impact if they can see the coarse number of affected individuals. 
The summary plot [@Potter2010] was also developed to address concerns about uncertainty in data displays. This dense display combines the plot of a single set of numeric values with a density estimate, boxplot, and statistical moments.

Some authors take this approach because they explicitly believe uncertainty is a variable of importance [@Blenkinsop2000], while others straddle the line, asserting uncertainty is acting as signal and noise, and should fulfil both roles [@geointerviews]. Simply put, these approaches might be described as swapping out the statistic for an "uncertainty statistic" to get an "uncertainty visualisation". Is it really that easy?

#### Example: visualising variance
@fig-statistic depicts the six plots showing this approach for the data introduced in @fig-ignore.  The original central value estimate has been replaced with an uncertainty statistic.
Plots R1, R2 are absurd examples of this approach: the regression line replaces the estimated slope and intercept of a line with their standard errors. 
To understand to plot you would read: the intercept of the line now says that the standard error of the trend is close to 0 in R1 and 20 in R2, and the error of the slope is low for R1 and very high for R2.
Sometimes this approach is sensible, as we can see in the exceedance probability maps depicted in plots S1 and S2. These map $P(temperature>27)$ to colour for each county. The sine wave trend is clearly visible when error is low but barely visible with high error. This does reveal something interesting: the variance is constant in G1 but different for each group in G2. It is a nonsensical display, though, because the trend has completely disappeared. In practice, two plots are typically presented: one showing the main estimate and the showing the uncertainty.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-statistic
#| fig-height: 6
#| fig-width: 10
#| out-width: 100%
#| fig-cap: "Treating the uncertainty as a statistic, to the extreme, with the same six examples. The easiest to digest is the choropleth (S1, S2): instead of temperature, the probability of exceeding 27$^o$C is shown. This has the effect of highlighting (sine wave) trend in the low error data, and de-emphasizing it in the high error data. The other four take some thinking. In R1, R2, the line now indicates the standard error of the intercept and slope of the regression line, and one would read that the error for 'R1 - low' is close to 0 but 'R2 - high' is closer to 20, and similarly for slope. Yes, this is an absurd display! With G1 and G2, it is similar, but we learn something interesting that was not seen earlier: the variance in G2 is not uniform like that in G1."

reg2a + exceed1 + uni2a + reg2b + exceed2 + uni2b + plot_layout(ncol=3, widths=c(30, 40, 30))  
```

#### What is an uncertainty visualisation, then?
What an uncertainty visualisation is or is not is one of the most pervasive divides in the literature.
For example, @Leland2005 mentions that popular graphics, such as pie charts and bar charts, omit uncertainty. @Wickham2011 suggests their product plot framework, for area plots like bar charts, and also histograms, needs to be extended to include uncertainty representation.
However, pie charts, bar charts and histograms have all been used in a significant number of experiments as examples of an "uncertainty visualisation" [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. 
What is going on here?

This conflict stems from a subconscious disagreement about the purpose of uncertainty visualisation.
If you believe uncertainty visualisation is about communicating risks or random variables, uncertainty visualisations are just visualisations of "uncertainty statistics" or distributions.
On the other hand, if you believe uncertainty visualisation is about suppressing false signals visually, then you see an uncertainty visualisation as a transformation of an existing graphic that adds the uncertainty in.
The former has no limitation on the visual appearance of an "uncertainty visualisation", allowing pie charts, bar charts or histograms, so long as the graphic is visualising "uncertainty", while the latter believes uncertainty visualisations only exist in relation to some "normal" visualisation.
When we refer to the graphics depicted in @fig-statistic as "uncertainty visualisations", we are classifying visualisations by the data they display, not their visual features.
This is not the standard approach in statistical graphics. 
A scatter plot that compares means and a scatter plot that compares variances are both scatter plots. 

Unlike plots, which are not defined by a statistic, uncertainty is *only* defined in relation to our visual statistic, a topic that frequently appears in the literature to be dependent on the "goals" of our analysis.
@Meng2014 commented that what is kept as data and what is tossed away is determined by the motivation of an analysis - what was previously noise can become signal depending on the question. @Otsuka2023 suggested that the process of observing data to calculate statistics is largely dependent on our goals, because the process of boiling real-world entities down into probabilities depends on the relationships we seek to identify within our data.
@Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision-maker wants to represent, and why it matters. 
@Fischhoff2014 suggested we should have methods for communicating uncertainty depending on what the user is supposed to do with it. @Spiegelhalter2017 says we "cannot assess the quality of risk communication unless the objectives are clear". 
@geointerviews asserted that whether or not it XXX uncertainty? XXX is a source of doubt depends on the context.
The sentiment behind this repeated point is clear: the role of uncertainty or signal is not dependent on the "type" of statistic, on the source of the information, or the methods we use; it is determined by the statistic we wish to draw inference on.
Therefore, the fundamental problem with the "uncertainty statistic" approach is that the uncertainty in the plot isn't acting as noise; it is acting as signal.

If the uncertainty in a graphic is acting as a signal, there isn't an interesting perceptual challenge associated with the visualisation: the uncertainty can be displayed using standard principles of graphic design.
In changing the inferential statistic, we also haven't dealt with the original problem of integrating noise, as these "uncertainty statistics" *also have associated uncertainty in the estimates* (e.g. variance of standard deviation estimate) that is being ignored.
There is nothing wrong with explicitly visualising variance, error, bias, or any other statistic. 
These metrics provide important and useful information for analysis and decisions.
The problem with this approach is that it means everything is an uncertainty visualisation, and if everything is an uncertainty visualisation, nothing is.

### Uncertainty as a variable
Another common characterisation of uncertainty is as just another variable to be integrated into the visualisation, which means uncertainty visualisation is, at its core, a high-dimensional visualisation problem [@Griethe2006; @geointerviews].
This occurs within computer science [@Kinkeldey2014], cartography [@MacEachren2005], and statistical graphics [@Leland2005].

Discussion of the visualisations focus on how "integrated" the uncertainty is with the estimate.
@Kinkeldey2014 identified a split between intrinsic plots, where we map uncertainty to the colour or size of the geometric object of our estimate, and extrinsic plots, where uncertainty is mapped to a separate geometric object, such as glyphs or error bars.
Similarly, @uncertchap2022 classified uncertainty visualisations as graphical annotations (extrinsic), and probability mapped to a visual encoding channel (intrinsic), or a hybrid of the two.
It is unclear if these levels of "integration" in a plot design affect its ability to suppress signals. 

#### Example: mapping two independent variables
@fig-variable shows the six examples introduced in @fig-ignore. Here, uncertainty is mapped to a spare aesthetic in the plot. In the grammar of graphics, variables are mapped to aesthetics, like position, colour, and size, within a plot.
In plots R1 and R2, the standard error of the slope is represented by the width of the line. This is common, but fails to represent the standard error of the intercept alongside the slope. We might think we can examine the width of the line at `wt=0`, but this would be incorrect. Here it has been included, less obviously, by mapping the standard error of the intercept to transparency. While these help to give a gestalt of the uncertainty, they are not exacting 
representations. One would expect that the width above and below the line is one standard error, but why not map the width to two standard deviations, or even three? Interpreting transparency into a numerical quantity is also virtually impossible, so using this aesthetic mapping is effectively useless.  

A bivariate colour palette map is shown in plots S1, S2. With two dimensions of the plot reserved for spatial position, it is difficult to incorporate the error. The bivariate colour palette maps the estimate to hue and the error to saturation, keeping the signal and noise contained to one visual aesthetic. This has the unfortunate effect of making the signal appear stronger when the error is higher (S2). Colour perception is a wild beast that is hard to tame. 

The extrinsic approach, shown in plots G1 and G2, has the estimate represented by a point, and uncertainty computed as a 95% confidence interval mapped to line length. The trend is still visible in both displays. It could be argued that where the variance is high, the trend remains a main focus; that is, the display fails to sufficiently suppress the signal.

Because all of these graphics visualise the distribution's estimate and uncertainty as two separate pieces of information, the message of the plot is  "here is the trend *and* here is the uncertainty".
It is worthwhile to examine why this occurs, to see if we can move towards a version of this plot where we are able to communicate signal and noise simultaneously.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-variable
#| fig-height: 6
#| fig-width: 10
#| out-width: 100%
#| fig-cap: "Treating the uncertainty as a variable, with the same six examples. In plots R1, R2 the standard error of the slope mapped to the line width. The standard error of the intercept is mapped to the transparency, which is less conventional. In plots S1, S2 a bivariate colour palette is used with mean mapped to the hue, and error mapped to saturation. Plots G1 and G2 represent the mean of each group as a point, and variance as an interval. The uncertainty is integrated well with the signal, but for the choropleth the result is undesirable: the signal is easier to see when the error is higher."

reg3a + bivar1 + uni3a + reg3b + bivar2 + uni3b + plot_layout(ncol=3, widths=c(30, 40, 30)) 

```

#### Can we visualise a "single integrated uncertain value"?
The reality is, based on our discussion on inferential statistics, uncertainty *isn't* a separate variable: it is a component of the random variable that is indistinguishable from the random variable itself.
Similarities between the "as a variable" approach and the "as a statistic" approach are apparent when we read motivations for visualising an estimate (i.e. @fig-ignore) and variance (i.e. @fig-statistic) side-by-side using two separate graphics. This organisation is vulnerable to change blindness [@simons1997] as one needs to switch focus between two displays.)
The preference for the methods utilised in @fig-variable is usually motivated by the difficulties in combining information on two separate graphics [@moritz2017trust; @Correll2018], rather than an understanding that the "uncertainty statistic" approach is not philosophically sound. 
To achieve signal suppression, we need to visualise noise and signal together as a “single integrated uncertain value” [@Kinkeldey2014] rather than as two separate statistics.

Just because we can still see the signal in @fig-variable, that does not mean the reading of the estimate is completely independent of the uncertainty.
When making any visualisations, we usually want the visual channels to be separable, that is, we don't want the data represented through one visual channel to interfere with the others [@Smart2019]. It is also interesting to know whether readers can see all structures when there are multiple structures present in a data plot, or whether they systematically fixate on one [@vanderplas2017].
Separability may be desireable in standard data visualisation, but in uncertainty visualisation it allows the estimate and its variance to be read independently, potentially leading to the uncertainty being ignored [@uncertchap2022]. 
Therefore, rather than trying to maintain visual separability, the goals of uncertainty visualisation align far better with the pursuit of visual integration.
In an ideal system, our estimate and uncertainty would be manipulated separately, but would be so *well-integrated that they are read as a single channel by the human brain*.
The problem is that even if we can implement the most extreme versions of integrable, our methods fall short, as illustrated by the bivariate colour palette map in @fig-variable. 
Colour hue and brightness are one of the classic examples of integrable variables [@Vanderplas2020], and decreasing saturation should make the colours harder to distinguish, but the signal is still clearly visible with the high variance. 
This is to say nothing of the fact that multi-dimensional colour palettes can make the graphics harder to read and less accessible [@Vanderplas2015].

#### Another example: mapping combined variables
The Value Suppressing Uncertainty Palette (VSUP) [@Correll2018] was designed with the intention of preventing high uncertainty values from being extracted from a map by blending colours together as they become less certain.
@fig-vsup shows the spatial example (S1, S2) using the VSUP approach.
Since the palette was designed with the extraction of individual values in mind and it has only been tested on simple value extraction tasks [@Correll2018] or search tasks [@Ndlovu2023], we can see that, at least for our example, when the uncertainty is high, the spatial trend has functionally disappeared. 

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-vsup
#| fig-height: 4.5
#| fig-width: 9
#| fig-cap: "The spatial examples displayed with a choropleth map using a VSUP colour palette, where hue is blended when increased uncertainty. Plot b has successfully produced signal suppression. Look closely at the scales, though: it may have you question whether this should have happened."

vsup1 | vsup2 

```

Finally, we have signal suppression!
Well, not really, sorry, we tricked you.
The two plots depicted in @fig-vsup actually show the exact same data; they are both the low variance case.
If you look closely, you can see that the two plots have a different scale, where plot (a) has been scaled according to our existing knowledge about this data, while plot (b) has been scaled using the range of the data passed to the plot. 
Since the variance and estimate are scaled independently, arbitrary differences in the range of our variance, unrelated to the estimate itself, will have significant impacts on the visual appearance of our plot.
The scale issue in VSUP maps was also recognised by @Kay2019, who noted that the suppression of any one hypothesis largely depends on the methods we use to combine the palette, and the variance levels at which the blending occurs.
This means that, for us to know that our plot will successfully perform signal suppression, we need to already know what signal we are trying to suppress and set up the VSUP palette accordingly.
This means that VSUP maps are not suitable for exploratory data analysis.

#### Uncertainty and exploratory data analysis
The lack of uncertainty in descriptive statistics is due to the lack of inference. 
<!--The reason there is no uncertainty in descriptive statistics is not due to a unique property of descriptive statistics, but rather due to the lack of inference.-->
Descriptive statistics are actually a small piece of a much larger field, exploratory data analysis (EDA), that tends not to perform statistical inference.
@Tukey1977 described EDA as the process of searching for interesting hypotheses ("the greatest value of a picture is when it forces us to notice what we never expected to see"), and defined it in relation to confirmatory data analysis (CDA), the process of verifying a hypothesis. There are more subfields of EDA: initial data analysis [@huebner2016; @chatfield] which involves checking assumption and data quality prior to CDA, and model diagnostics (e.g. @bkw1980), including posterior checks of model fit. What binds these pursuits together is their reliance on visual summaries for making assessments, and an absence of formal inference.  

 @Hullman2021 argued that the EDA and CDA are not entirely distinct, as it is often difficult to draw a hard line. 
Our belief, as with many concepts, is that these approaches exist on a continuum, where we have an inherent trade-off between the number of hypotheses we can look for and the certainty of any conclusions reached. 
It can help to think of the knowledge-generating process of EDA and CDA as the nozzle on a hose with multiple spray options, where EDA is a fine misting spray that touches everything in the room, and CDA is a high-pressure jet capable of obliterating any and all debris from any single spot. 

Viewing EDA and CDA as a dichotomy can create some confusion when it comes to understanding the source of uncertainty in our analysis.
This is why we have avoided the topic until now, despite the fact that an uncertainty visualisation system for EDA is one of the most discussed topics in the field [@Hadjimichael2024; @geointerviews; @MacEachren2005; @Sarma2024; @Griethe2006].
The EDA versus CDA dichotomy can be compared to the dichotomy between induction, for building theories, and deduction, for testing theories, from formal logic.
One of the trade-offs in the two methods is that deductive conclusions provide certainty, while conclusions from induction are inherently uncertain. 
This means that EDA, the inductive counterpart, has uncertainty in any conclusions reached, with the requirement to follow up our newfound hypothesis with CDA if we want true certainty. 
This adds another layer of confusion to the study of uncertainty visualisation: conflating the uncertainty in our data with the uncertainty that is inherent to an exploratory process (EDA). 

Authors often over-compensate for the inherent EDA uncertainty, pre-emptively hedging against every false inference that could possibly be drawn from a graphic. @Hullman2021 argues there is no such thing as a "model-free" visualisation. @Guo2024 provides a grammar for visualising statistical model checks. The lineup protocol [@Buja2009] provide the viewer with plots of the data in a field of plots of null data where any patterns seen are due to sampling variability. The Rorschach protocol, from the same paper, shows only null plots to give the reader some intuition for what spurious sampling patterns exists. @Savvides2019 provides a statistical super-test against multiple comparisons driven by probabilistic arguments. There is a CDA quality to these approaches. The sum total of a lot of CDA is not EDA, just as swinging a high-pressure jet around a room is not equivalent to using a misting spray. While EDA and CDA may be along a continuum, 
we cannot simultaneously perform EDA and CDA as the approaches are, philosophically speaking, perpendicular to one another.

To truly create an uncertainty visualisation approach that is capable of EDA, we need to accept that uncertainty is inherent to the method, and it cannot be pre-emptively removed from the visualisations. 
If we accept this fact, then the only possible source of uncertainty in an uncertainty visualisation system that performs EDA is from the data itself. 
That is to say, the only possible way for there to be uncertainty in a visualisation designed for EDA is that the data itself represents inference that has been done earlier in our analysis.
We can see this in the original description of @fig-ignore where the uncertainty in all cases, even measurement error, represents inference that was performed earlier in our analysis.

In the VSUP approach, our ability to arbitrarily decide which values to blend at, or which suppression approach to use, means that the "uncertainty" we are visualising will be informed by the conclusions we are drawing, and not a product of the data itself -- an antithetical approach to EDA.
For a visualisation to be suitable for EDA, it should always look the same regardless of what hypothesis we plan to draw from it.
As much of the transparency in data visualisation comes from this feature in EDA, it is reasonable to set it as a requirement of our visualisations.
Ensuring this property means we cannot treat signal and noise as separate variables, but rather as a single integrated unit.

### Uncertainty as a distribution
Rather than trying to reduce random variables to a single value or pair of values, why not visualise the whole distribution?
This approach is found in computer science's hypothetical outcome plots (HOPs) which animate a sequence of potential outcomes of a distribution [@Hullman2015], in geoscience's pixel-maps [@Lucchesi2021; @Blenkinsop2000], and in statistics multiple forecasts [@fpp3].

#### Example: visualising samples
@fig-variable shows the six examples introduced in @fig-ignore, shown as distributions. The distribution is represented by either a sample of outcomes or a quantile dot plot.
Plots R1 and R2 show a linear regression as a sample of possible outcomes from the distribution. The distribution used for both the slope and intercept is normal with the conventional mean and standard error.
Plots S1 and S2 show a pixel map, which is a choropleth map where each county is coloured by a sample of outcomes from the distribution.
Plots G1 and G2 show each univariate distribution as a quantile dot plot.
We can see that the strong downward trend in the linear regression, the sine wave in the choropleth map, and the incremental increase in the univariate distributions are all clearly visible in the low variance case, but disappear in the high variance cases. 
The graphics have achieved signal suppression.
Visualising the random variable as a distribution gives additional information, such as the previously hidden bimodality of the univariate distributions in G1 and G2.


```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-distribution
#| fig-height: 6
#| fig-width: 10
#| out-width: 100%
#| fig-cap: "Treating the uncertainty as a distribution, with the same six examples. Plots R1 and R2 show a linear regression as a sample of possible outcomes from the distribution. Plots S1 and S2 show a pixel map. Plots G1 and G2 show the groups as quantile dot plots. In each case the signal (regression line, sine wave, increasing trend) has disappeared with high uncertainty."

reg4a + pixel1 + uni4a + reg4b + pixel2 + uni4b + plot_layout(ncol=3, widths=c(30, 40, 30)) 


```

#### Quantified versus unquantified uncertainty
By showing our distribution as "data", we are able to read the "uncertainty" plots using the same perceptual mechanisms we use to read the  "non-uncertainty" plot.
This should lead to more effective communication, as people tend to read more complicated visualisations like the bivariate and VSUP plots the same way they read the simple choropleth counterpart [@Ndlovu2023].
This approach also does not significantly hinder our ability to extract the individual values mapped by the previous plots, as extracting global statistics from a sample can be done with relative ease [@Franconeri2021]. 
We are able to include more information by offloading more computation to visual processing, but what is the limitation on this approach?
Which aspects of uncertainty should be processed by our visual system, and which should be processed by our statistical computation?
It is not obvious from the question, but this is actually a question about how much of our uncertainty should be quantified. 

Quantified uncertainty usually focuses narrowly on concepts such as probability, confidence intervals, variance, error, or precision [@Hullman2018; @Maceachren2012; @Thomson2005], while unquantified uncertainty often includes a broader range of concepts like missing values, reliability, model validity, or source integrity [@Griethe2006; @Leland2005; @Pang1997; @Pham2009; @Boukhelifa2017]. 
When discussing uncertainty, we typically include these unquantified uncertainties, not because these things *are* uncertainty, but because they can *create* uncertainty when we perform inference.
This is often because these unquantified uncertainties violate our assumptions of the uniformity of nature [@Otsuka2023].

Sometimes we are able to visualise these assumption violations directly.
For example, we can check for structure in our missing data using the `naniar` package @Tierney2023 that allows us to include missing values as a "shadow" alongside our usual visualisations. 
This approach amounts to just "showing the data", which is a simple but effective option for uncertainty visualisation that is largely overlooked.
While this approach is useful for better understanding data, it will not eliminate trends that have become invalid due to structure in missing data or an invalid model.
We can only integrate uncertainty as noise when that uncertainty has been *quantified* as an effect on the estimates we are visualising. 
This is not to say one method is preferable; visualising both quantified and unquantified uncertainty is necessary for a healthy analysis.
Data analysis often works in cycles, where we find assumption violations using EDA, quantify the effect of these violations on inference, and then visualise the output of that inference using uncertainty visualisation.

## Evaluating uncertainty visualisations
Unfortunately, the conflicting results in the field are not limited to plot design and extend to the experimental findings as well [@MacEachren2005; @Kinkeldey2014; @Hullman2016].
There are as many explanations for the noisy evaluation studies as there are contradictions in the research itself. 
@Kim2019 believes there is some interference in results from participants' prior beliefs; 
@Hullman2016 believes the noise in the literature could come from visual heuristics, subjective probabilities, unknown participant utility functions, or a misunderstanding of statistical concepts (such as confidence intervals).
@Kinkeldey2014 suggest the perception of visualisation changes by audience, so we cannot expect the same results between different subpopulations. 
@Brennen2018 attributes evaluation difficulties to cognitive load from complicated uncertainty visualisations, as well as the participants' prior experience in the topic.
While these issues will certainly have some impact on our ability to synthesise, none of them are unique to uncertainty visualisation. 
Rather, the issue is likely due to a disconnect between the evaluation methods used and the stated goals each experiment, a common issue in visualisation evaluation experiments [@Vanderplas2020].


### Current evaluation methods

#### Value extraction
Uncertainty visualisations are most commonly evaluated based on how accurately viewers can extract an estimate and its variance [@Kinkeldey2014; @Hullman2019]. 
This is not unusual, as direct observation is the simplest way to verify that information can be accurately read from a graph [@Vanderplas2020].
Unfortunately, this approach doesn't work for uncertainty visualisation.
The second we ask a specific question about a statistic, that statistic becomes inferential, even if the plot was not the intent behind the question. 
By shifting the focus from $\hat{X}$ to $Var(\hat{X})$ or $P(\hat{X}<x)$, we end up evaluating visualisations on their ability to convey uncertainty statistics, rather than their ability to perform signal suppression.
Even if the authors do not realise it themselves, there is nothing unique to uncertainty in these studies, so when we boil the findings down to generalised results, they simply restate existing principles within information visualisation.
Some of the findings are obvious: participants were more accurate when reading a probability expressed as text than when they had to extract it from a graphic [@Cheong2016; @Savelli2013].
Other studies replicate existing research, such as the finding that a probability mapped to a position is more accurate than one mapped to an area [@Ibrekk1987; @Gschwandtnei2016], established  as part of the hierarchy of perceptual tasks more than 40 years ago [@Cleveland1984; replicated by @heer2010]. 
This extends beyond simple accuracy evaluations: @Sanyal2009 found that colour was more effective than size when searching for extrema in variances; we have known that pre-attentive aesthetics, such as colour, are more efficient for search tasks since the 1980s [@Vanderplas2020].
By classifying these studies as evaluations of "uncertainty visualisation" while evaluating uncertainty as a signal, we are encouraged to see successful examples of signal suppression as failure. 
This approach leads authors to advise against particular aesthetic mappings for uncertainty, because they cause participants to have more difficulty extracting values [@Blenkinsop2000].
This conclusion is antithetical to the goals of signal suppression and occurs because these methods evaluate uncertainty as a signal, not as noise. 

#### Trust, confidence, and risk aversion
If we cannot directly measure uncertainty for fear that it turns into a signal, we might then assume we can measure the secondary benefits of increased transparency.
This seems to be the approach of many visualisation authors, as secondary benefits such as trust, confidence, and risk aversion are all frequently used in uncertainty evaluation studies [@Hullman2019].
Unfortunately, measuring these secondary effects often leads to confusing conclusions that simultaneously argue for and against the inclusion of uncertainty.

This is most commonly noticed in the use of trust as a measure, as several authors have commented that measuring trust, and not transparency, can lead to a questionable subtext that argues against transparency [@Spiegelhalter2017; @ONeill2018].
We see this directly play out in the visualisation literature, where surveyed visualisation authors explicitly said they didn't include uncertainty due to the fact that they might decrease trust in their conclusions [@Hullman2020a].
This sentiment is also true for confidence, as @Blenkinsop2000 commented that visually integrable depictions of uncertainty should be avoided, as they decrease the viewer's confidence in their extracted data values.
Confidence is also used as a proxy for poorly constructed visualisation in standard visualisation experiments, so using confidence as a measure of uncertainty collapses these confounding factors down into a single metric. 
<!--XXX citation? We use it as a measure but it doesn't correlate with much usually XXX-->

Another secondary effect that is similar to trust and confidence is risk aversion.
Risk aversion is an economic term used to describe an agent who would choose a random variable with a lower expected payout because it also has a lower variance.
Risk aversion's role in the uncertainty visualisation is unclear, as authors will argue uncertainty should elicit more risk aversion in one paper [@Hullman2019], and argue for less risk aversion (by proxy of suggesting rational agents as a benchmark) in the next [@wu2023rational].
Ultimately, these approaches have similar issues to value extraction studies, except they are slightly more confusing in their goals, leading them to simultaneously argue for and against the inclusion of uncertainty in a visualisation.

#### Alternative approaches
Often, authors understand that the effects of uncertainty are more complicated than simple value extraction.
While these approaches are a step in the right direction,  evaluating uncertainty or visualisations of distributions without a solid definition of the effects can lead to unintended consequences, no matter what approach we take.

One approach is to ask open-ended questions, such as asking participants for the "best estimate" [@Ibrekk1987], or to select which distribution is the "furthest to the right" from a lineup [@Hofmann2012]. 
<!-- I agree with Di, this isn't open-ended -- it's a multiple choice question with 20 possible responses -->
In both cases, the ground truth is based on the mean of the distribution.
This approach can lead to inconclusive results, as we are left unclear if it was the phrasing of the question or the plot design that caused the participants to answer incorrectly.

On the other hand, questions that are incredibly specific about the distribution information can confuse the participants and induce noisy results.
For example,  @Hullman2015 asked participants to compare two normally distributed groups, A and B, and had many participants say that group A was more likely to be bigger, despite group B having a higher mean.
@Gschwandtnei2016 asked participants the "probability that the interval has already ended at the marked point in time?” and participants replied with the probability that the interval had already started.

The confusion around trying to capture the effects of uncertainty can also (understandably) extend to the authors of the study itself. 
We can see an example of this in @Padilla2017, In order to answer the question correctly, the first experiment required participants to assume that an oil rig being "more likely to be hit" by a hurricane would *not* translate to the rig sustaining "more damage". The second experiment required participants to assume the opposite.

#### Effective methods
This is not to say all evaluation studies fail to properly evaluate uncertainty as noise. 
There are several studies that ask participants to identify a particular signal that the noise is trying to obfuscate [@Kale2018; @Correll2014], which seems to be an effective method. 
The only problem with these studies is that most uncertainty visualisation methods are not integrated into *the grammar of graphics* [@Leland2005; @ggplot2] so we regularly see comparisons between disparate plots that would never be considered substitutes for one another outside the artificial "uncertainty visualisation" framework they are placed within.
For example, @Kale2018 compared static bar charts with error bars to a bar chart with animated samples, meaning that any difference in participants' ability to read the plot could be due to the statistic (confidence interval versus sample), the geometry (bars versus intervals), or the use of animation (static versus animated plots).
This means that even when evaluations *are* done correctly, there is no generalisable theory we can take from the results.

The point here is not to accuse the authors of poor academic rigour.
The papers are (usually) logically consistent and well-formulated pieces of work.
Rather, the point is to illustrate that evaluating uncertainty as noise is surprisingly difficult.
Designing tests for signal suppression will require a formalisation of uncertainty within the grammar of graphics, as well as improved evaluation methods.

### Implicit Hypothesis Testing
The main problem with current uncertainty visualisation evaluations is that they often require explicit (or convoluted) questions about the variance.
Asking direct questions about the statistics or outcomes is not an explicit requirement of visualisation evaluations.
In their review of testing statistical graphics, @Vanderplas2020 drew a distinction between explicit tests, where participants are asked direct questions about specific features of a plot, and implicit testing, where users identify both the purpose and function of the plot.
The lineup protocol is the most salient example of the implicit approach.
Lineups are a confirmatory visualisation tool where participants are shown a set of $M$ plots, and asked to identify the plot that is the "most different", leaving participants to decide what "most different" means to them, even if it is not what the authors intended [@vanderplas2017].
The implicit test does not limit the versatility of the approach, with the lineup being used to evaluate the effectiveness of different types of plots [@Hofmann2012], colour palettes [@Reda2021], and design decisions [@vanderplas2017].

Lineup protocols are not only useful for implicit testing: they also have parallels to hypothesis testing that can be leveraged in uncertainty visualisation.
The concept of signal suppression is, at its core, an assertion of statistical validity: the visibility of signals should be directly proportional to $p$-values or some equivalent measure.
This comparison is not new in uncertainty visualisation, where parallels have been drawn to frequentist statistics by @Correll2014, who compared results to Cohen's D, and to Bayesian statistics by @Kim2019, who evaluated plots based on their impact on the users' prior beliefs. 
The comparison to hypothesis testing is far more natural for the lineup protocol, which has a visual test statistic [@Majumder2013] and can be compared to standard statistical tests using power curves [@Patrick2023]. 
The connections between lineups and uncertainty visualisation are numerous and have been previously identified in the development of HOPs [@Hullman2015].

The lineup protocol and uncertainty visualisations are similar: lineups were designed for checking if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010].
As both approaches are attempting to do the same thing, it is likely that we are unable to leverage the lineup protocol directly to evaluate uncertainty visualisation, but a new evaluation methodology should try to learn from the success of the lineup approach. 
Designing an implicit testing method for uncertainty visualisation that allows us to draw parallels to standard notions of statistical significance would solve many of the issues with the current evaluation approaches.


## Conclusions and Future Work
This paper examines the literature and provides suggestions for a structural framework to supports uncertainty visualisation. Particularly, we propose that uncertainty visualisation should accomplish signal-suppression, dampening weak signals and amplifying strong signals. We have also highlighted several gaps in the existing literature.

*Experimental practices on uncertainty visualisation need standards.* 
Some existing evaluation experiments treat uncertainty as a signal, while others treat uncertainty as noise. 
As a result, it is difficult to combine results from papers to get a meaningful sense of how uncertainty information is understood by a viewer.
Researchers need to ensure that when they identify the motivation behind their visualisation technique, their evaluation methods align with the stated goals of the paper. 

*Experimental methods that evaluate uncertainty as noise need to be developed.*
Research into separability and integrability of signal and noise is of particular interest to uncertainty visualisation, as it allows assessment of the interference between the two. 
When designing experiments, authors often choose aesthetics that are visually distinguishable; uncertainty visualisation authors should consider doing the opposite. 

*Uncertainty needs to be formalised within the grammar of graphics.* 
Some of this formalisation was done by @Kay2023, but it focuses only on the visualisation of univariate distributions.
Giving authors the ability to describe uncertainty visualisations in terms of statistics, geometries and aesthetics will support evaluation experiments that can build towards a cohesive theory of visualising uncertainty.

*Software that allows users to easily perform signal suppression is needed.* 
Existing uncertainty visualisation methods view a distribution as its own object, and there are no software options treating "an uncertainty visualisation as a function of an existing visualisation" philosophy.  

Signal suppression is an undeveloped area of visualisation research, and developing methods for the practice may require us to challenge our entire notion of what makes a good visualisation.

## Acknowledgements

The first author of this paper is supported in part by a scholarship from the Australian Energy Market Operator.
This research was supported by the Commonwealth through an Australian Government Research Training Program Scholarship [DOI: https://doi.org/10.82133/C42F-K220]. 
The R packages were used for this work were: `tidyverse` [@tidyverse], `RColorBrewer` [@RColorBrewer], `scales` [@scales], `sf` [@sf], `urbnmapr` [@urbnmapr], `flextable` [@flextable], `colorspace` [@colorspace], `ggdist` [@Kay2023], `ggdibbler` [@ggdibbler], `patchwork` [@patchwork], `distributional` [@distributional], `ggthemes` [@ggthemes], and `rgeos` [@rgeos]. 
The GitHub repository for this paper can be found at https://github.com/harriet-mason/ARSA-UncertaintyLitReview, which contains the files required to reproduce this article in full.

## Bibliography

```{r, include=FALSE, eval=FALSE}
library(spelling)
qmd <- "paper.qmd"
ignore <- readLines("WORDLIST")
check_spelling <- spell_check_files(
  qmd,
  ignore = ignore,
  lang = "en_GB"
)
if (nrow(check_spelling) > 0) {
  print(check_spelling)
  stop("Check spelling in Qmd files!")
}
```






